Screening Sinkhorn Algorithm via Dual Projections

### 2019-05-13 tex/v2/screenkhorn.tex
Add first draft of paper

This paper deals with the problem of approximating optimal transport (OT) distance between two discrete measures.
Our proposed approach involves a convex projection of the \emph{dual of Sinkhorn divergence}, allowing to define two appropriate active indices sets for the potential variables.
These indices sets depend on two parameters acting like a threshold and a scaling factor and they are both directly linked to a priori fixed number budget of points from the supports of the given discrete measures.
This new analysis induces a screened version of the dual of Sinkhorn divergence and suggests the \emph{Screenkhorn} algorithm.
We illustrate the favorable performance of Screenkhorn in practice with numerical experiments on synthetic and real datasets.
\end{abstract}

@@

Computing OT distances between pairs of probability measures or histograms, such as the earth mover's distance~\cite{werman1985,Rubner2000} and Monge-Kantorovich or Wasserstein distance~\cite{villani09optimal}, are currently generating an increasing attraction in different machine learning tasks~\cite{pmlr-v32-solomon14,kusnerb2015,pmlr-v70-arjovsky17a,ho2017}, statistics~\cite{frogner2015nips,panaretos2016,ebert2017ConstructionON,bigot2017,flamary2018WDA}, and computer vision~\cite{bonnel2011,Rubner2000,solomon2015}, among other applications~\cite{klouri17,peyre2019COTnowpublisher}.
In many of these problems, OT exploits the geometric features of the objects at hand in the underlying spaces to be leveraged in comparing probability measures.
This effectively leads to improve performance of methods that are oblivious to the geometry, for example the chi-squared distances or the Kullback-Leibler divergence.
Unfortunately, this advantage comes at the price of an enormous computational cost of solving the OT problem, that can be prohibitive in large scale applications.
For instance, the OT between two histograms with supports of equal size $n$ can be formulated as a linear programming problem that requires generally $\bigO(n^3\log n)$~\cite{pele2009} arithmetic operations, which is problematic when $n$ is larger than $10^3.$

@@

We give a new algorithm to approximate the regularized OT distance between discrete measures. 
Our algorithmic analysis is based on an approximate of the dual of Sinkhorn divergence by adding new constraints feasibility. 
These constraints are defined through a convex set which depends on two parameters, acting like threshold and scaling factor.
We prove that dual solution of this approximation guarantees the existence of two active indices sets for the potential variables.
These active sets are both directly linked to a priori fixed number budget of points from the supports of the given discrete measures.
We then restrict the constraints feasibility with respect to the active sets to get a ``screened'' version of the dual of Sinkhorn divergence, and hence we develop the Screenkhorn algorithm.
% The proposed algorithm consits of

@@

Below we provide the derivation of the dual problem for the regularized OT problem~\eqref{sinkhorn-primal}. Towards this end, we begin with writing its Lagrangian dual function :
\begin{equation*}
  \mathscr{L}(P,y, z) = \inr{C,P} + \eta \inr{\log P, P} + \inr{y, P\mathbf{1}_m - \mu} + \inr{z,P^\top \mathbf{1}_n - \nu}.
\end{equation*}
The dual of Sinkhorn divergence can be derived by solving $\min_{P \in \R_+^{n\times m}}\mathscr{L}(P,y, z)$. It is easy to check that objective function $P\mapsto \mathscr{L}(P,y, z)$ is strongly convex and differentiable. Hence, one can solve the latter minimum by setting $\nabla_P \mathscr{L}(P,y, z)$ to $\mathbf{0}_{n\times m}$. Therefore, we get 
\begin{equation}
  P^\star_{ij} = \exp\Big(- \frac{1}{\eta} (y_i + z_j + C_{ij}) - 1\Big), 
\end{equation}
for all $i=1, \ldots, n$ and $j=1, \ldots, m$. Plugging this solution,  and setting the change of variables $u = -y/\eta - 1/2$ and $v = - z/\eta - 1/2$, the dual problem is given by
\begin{equation}
\label{sinkhorn-dual}
\min_{u \in \R^n, v\in\R^m}\big\{\Psi(u,v):= \mathbf{1}_n^\top B(u,v)\mathbf{1}_m - \inr{u, \mu} - \inr{v, \nu} \big\},
\end{equation}
where $B(u,v) = \Delta(e^{u}) K \Delta(e^{v})$ and $K = e^{-C/\eta}$ stands for the Gibbs kernel associated to the cost matrix $C$. 
We refer to problem~\eqref{sinkhorn-dual} to the \emph{dual of Sinkhorn divergence}.
Therefore, the optimal solution of Sinkhorn divergence is given by $P^\star = \Delta(e^{u^\star}) K \Delta(e^{v^\star})$
where the couple $(u^\star, v^\star)$ satisfies:
\begin{align*}
\label{sinkhorn-dual}
  (u^\star, v^\star) &= \argmin_{u \in \R^{n}, v\in \R^m} \{\Psi(u,v)\}.
\end{align*}
Note that the matrices $\Delta(e^{u^\star})$ and $\Delta(e^{v^\star})$ are unique up to a constant factor~\citep{sinkhorn1967}.

@@

The Screenkhorn algorithm, presented in Algorithm~\ref{screenkhorn}, consists of two steps: the first one is an initialization where we calculate the active sets $I_{\varepsilon,\kappa}$, $J_{\varepsilon,\kappa}$. 
The second is a constrained L-BFGS solver~\cite{zhu1997-LBFGS} for the stacked vector $\theta=(u_{I_{\varepsilon,\kappa}},v_{J_{\varepsilon,\kappa}}) \in \R^{n_b\times m_b}.$ 
It is worth to note that the couple variables $(u,v)$ to be optimized in Screenkhorn belongs to $\R^{n_b\times m_b}$. Furthermore, it Screenkhorn uses only the restricted parts $K_{(I_{\varepsilon,\kappa}, J_{\varepsilon,\kappa})},$ $K_{(I_{\varepsilon,\kappa}, J^\complement_{\varepsilon,\kappa})},$ and $K_{(I^\complement_{\varepsilon,\kappa}, J_{\varepsilon,\kappa})},$ from the Gibbs matrix $K$, in contrast to Sinkhorn that performs alternating updates of all rows and columns of $K.$

@@

\LinesNotNumbered
\begin{algorithm}[htbp]
\SetNlSty{textbf}{}{.}
\DontPrintSemicolon
\caption{Screenkhorn}
\label{screenkhorn}
\textbf{input: }{$C$, $\eta$, $\mu \in \Sigma_n$, $\nu \in \Sigma_m$, $n_b$ and $m_b$;}\\

@@

\begin{lemma}
\label{lemma_bounds_on_marginals}
Let $(u^{\text{sc}}, v^{\text{sc}})$ be an optimal solution of problem~\eqref{screen-sinkhorn_second_def}.
Then one has 
\begin{equation*}
\norm{\mu^{\text{sc}}}_1 \leq \kappa \norm{\mu_{I_{\varepsilon,\kappa}}^{\text{}}}_1 + (n-n_b) \Big(\frac{m_b\max_{j \in J_{\varepsilon,\kappa}} \nu_j}{n\kappa\min_{i,j}K_{ij}} + (m-m_b)\varepsilon^2 \Big)
\end{equation*}
and 
\begin{equation*}
\norm{\nu^{\text{sc}}}_1 \leq \kappa^{-1} \norm{\nu_{J_{\varepsilon,\kappa}}}_1 + (m-m_b)\Big(\frac{n_b\kappa \max_{i\in I_{\varepsilon,\kappa}}\mu_i}{m\min_{i,j}K_{ij}} + (n-n_b)\varepsilon^2\Big).
\end{equation*}
\end{lemma}
The following Proposition gives also an upper bound of the marginal errors.
\begin{proposition}
\label{proposition_error_in_marginals}
One has 
\begin{align*}
\norm{{\mu} -{\mu}^{\text{sc}}}^2_1 \leq &7n_b(\kappa-\log(\kappa)-1)\max_{i} \mu_i + 7(n-n_b)\bigg(\frac{m_b\max_{j}\nu_j}{n\kappa\min_{i,j} K_{ij}} + (m- m_b) \varepsilon^2 - \min_{i}\mu_i\\
&+ \max_{i} \mu_i\log\Big(\frac{\kappa(n-n_b+ 1)\max_{i} \mu_i}{m_b\min_{i,j}K_{ij}\min_{j \in J_{\varepsilon,\kappa}}\nu_j} + \frac{\kappa^2\max_{i} \mu_i}{mm_b\varepsilon^2(\min_{i,j}K_{ij})^2\min_{j \in J_{\varepsilon,\kappa}}\nu_j}\Big)% \bigg\}^{1/2}
\end{align*}
and 
\begin{align*}
\norm{{\nu} -{\nu}^{\text{sc}}}^2_1 \leq &7m_b(\kappa-\log(\kappa)-1)\max_{j} \nu_j + 7(m-m_b)\bigg(\frac{n_b\kappa\max_{i}\mu_i}{n\min_{i,j} K_{ij}} + (n- n_b) \varepsilon^2 - \min_{j}\nu_j\\
&+ \max_{j} \nu_j\log\Big(\frac{\kappa(m-m_b+ 1)\max_{j} \nu_j}{n_b\min_{i,j}K_{ij}\min_{i \in I_{\varepsilon,\kappa}}\mu_i} + \frac{\kappa^2\max_{j} \nu_j}{nn_b\varepsilon^2(\min_{i,j}K_{ij})^2\min_{i \in I_{\varepsilon,\kappa}}\mu_i}\Big)% \bigg\}^{1/2}
\end{align*}

>>>>

This paper deals with the problem of approximating optimal transport (OT) distance between two discrete measures.
Our proposed approach involves a convex projection of the \emph{dual of Sinkhorn divergence}, allowing to formulate two appropriate active indices sets for the potential variables.
These indices sets depend on two parameters acting like a threshold and a scaling factor and they are both directly linked to a priori fixed number budget of points from the supports of the given discrete measures.
This new analysis induces a screened version of the dual of Sinkhorn divergence and suggests the \emph{Screenkhorn} algorithm.
We illustrate the favorable performance of Screenkhorn in practice with numerical experiments on synthetic and real datasets.
\end{abstract}

@@

Computing OT distances between pairs of probability measures or histograms, such as the earth mover's distance~\cite{werman1985,Rubner2000} and Monge-Kantorovich or Wasserstein distance~\cite{villani09optimal}, are currently generating an increasing attraction in different machine learning tasks~\cite{pmlr-v32-solomon14,kusnerb2015,pmlr-v70-arjovsky17a,ho2017}, statistics~\cite{frogner2015nips,panaretos2016,ebert2017ConstructionON,bigot2017,flamary2018WDA}, and computer vision~\cite{bonnel2011,Rubner2000,solomon2015}, among other applications~\cite{klouri17,peyre2019COTnowpublisher}.
In many of these problems, OT exploits the geometric features of the objects at hand in the underlying spaces to be leveraged in comparing probability measures.
This effectively leads to improve performance of methods that are oblivious to the geometry, for example the chi-squared distances or the Kullback-Leibler divergence.
Unfortunately, this advantage comes at the price of an enormous computational cost of solving the OT problem, that can be prohibitive in large scale applications.
For instance, the OT between two histograms with supports of equal size $n$ can be formulated as a linear programming problem that requires generally super $\bigO(n^3)$~\cite{pele2009} arithmetic operations, which is problematic when $n$ is larger than $10^3.$

@@

We give a new algorithm to approximate the regularized OT distance between discrete measures. 
Our algorithmic analysis is based on an approximate of the dual of Sinkhorn divergence by adding new constraints feasibility. 
These constraints are defined through a convex set which depends on two parameters, acting like threshold and scaling factor.
We prove that dual solution of this approximation guarantees the existence of two active indices sets for the potential variables.
These active sets are both directly linked to a priori fixed number budget of points from the supports of the given discrete measures.
We then restrict the constraints feasibility with respect to the active sets to get a ``screened'' version of the dual of Sinkhorn divergence. 
The Screenkhorn algorithm developed in this paper relies on two steps; the first one consists of an initialization step devoted to determine the active sets while the second is a constrained L-BFGS solver~\cite{zhu1997-LBFGS,cuturi2016SIAM,blondel2018ICML}.

@@

Below we provide the derivation of the dual problem for the regularized OT problem~\eqref{sinkhorn-primal}. Towards this end, we begin with writing its Lagrangian dual function :
\begin{equation*}
  \mathscr{L}(P,y, z) = \inr{C,P} + \eta \inr{\log P, P} + \inr{y, P\mathbf{1}_m - \mu} + \inr{z,P^\top \mathbf{1}_n - \nu}.
\end{equation*}
The dual of Sinkhorn divergence can be derived by solving $\min_{P \in \R_+^{n\times m}}\mathscr{L}(P,y, z)$. It is easy to check that objective function $P\mapsto \mathscr{L}(P,y, z)$ is strongly convex and differentiable. Hence, one can solve the latter minimum by setting $\nabla_P \mathscr{L}(P,y, z)$ to $\mathbf{0}_{n\times m}$. Therefore, we get 
\begin{equation}
  P^\star_{ij} = \exp\Big(- \frac{1}{\eta} (y_i + z_j + C_{ij}) - 1\Big), 
\end{equation}
for all $i=1, \ldots, n$ and $j=1, \ldots, m$. Plugging this solution,  and setting the change of variables $u = -y/\eta - 1/2$ and $v = - z/\eta - 1/2$, the dual problem is given by
\begin{equation}
\label{sinkhorn-dual}
\min_{u \in \R^n, v\in\R^m}\big\{\Psi(u,v):= \mathbf{1}_n^\top B(u,v)\mathbf{1}_m - \inr{u, \mu} - \inr{v, \nu} \big\},
\end{equation}
where $B(u,v) = \Delta(e^{u}) K \Delta(e^{v})$ and $K = e^{-C/\eta}$ stands for the Gibbs kernel associated to the cost matrix $C$. 
We refer to problem~\eqref{sinkhorn-dual} to the \emph{dual of Sinkhorn divergence}.
Therefore, the optimal solution $P^\star$ of Sinkhorn divergence takes the form $P^\star = \Delta(e^{u^\star}) K \Delta(e^{v^\star})$
where the couple $(u^\star, v^\star)$ satisfies:
\begin{align*}
\label{sinkhorn-dual}
  (u^\star, v^\star) &= \argmin_{u \in \R^{n}, v\in \R^m} \{\Psi(u,v)\}.
\end{align*}
Note that the matrices $\Delta(e^{u^\star})$ and $\Delta(e^{v^\star})$ are unique up to a constant factor~\citep{sinkhorn1967}. Furthermore, $P^\star$ can be solved efficiently by iterative Bregman projections~\cite{benamou2015IterativeBP} referred as Sinkhorn iterations, and the method is referred as Sinkhorn algorithm which, recently, is proven to achieve a near-$\bigO(n^2)$ complexity.

@@

Pseudocode of our proposed algorithm is given in Algorithm~\ref{screenkhorn}.
Screenkhorn consists of two steps: the first one is an initialization where we calculate the active sets $I_{\varepsilon,\kappa}$, $J_{\varepsilon,\kappa}$. 
The second is a constrained L-BFGS solver~\cite{zhu1997-LBFGS} for the stacked variable $\theta=(u_{I_{\varepsilon,\kappa}},v_{J_{\varepsilon,\kappa}}).$ 
% It is worth to note that the couple variables $(u,v)$ to be optimized in Screenkhorn belongs to $\R^{n_b\times m_b}$. 
It is worth to note that Screenkhorn uses only the restricted parts $K_{(I_{\varepsilon,\kappa}, J_{\varepsilon,\kappa})},$ $K_{(I_{\varepsilon,\kappa}, J^\complement_{\varepsilon,\kappa})},$ and $K_{(I^\complement_{\varepsilon,\kappa}, J_{\varepsilon,\kappa})}$ of the Gibbs kernel $K$, in contrast to Sinkhorn algorithm which performs alternating updates of all rows and columns of $K.$

@@

\LinesNotNumbered
\begin{algorithm}[htbp]
\SetNlSty{textbf}{}{.}
\DontPrintSemicolon
\caption{Screenkhorn$(C,\eta,\mu,\nu,n_b,m_b)$}
\label{screenkhorn}
% \textbf{input: }{$C$, $\eta$, $\mu \in \Sigma_n$, $\nu \in \Sigma_m$, $n_b$ and $m_b$;}\\

@@

\begin{lemma}
\label{lemma_bounds_on_marginals}
Let $(u^{\text{sc}}, v^{\text{sc}})$ be an optimal solution of problem~\eqref{screen-sinkhorn_second_def}.
Then one has 
\begin{equation*}
\norm{\mu^{\text{sc}}}_1 \leq \kappa \norm{\mu_{I_{\varepsilon,\kappa}}^{\text{}}}_1 + (n-n_b) \Big(\frac{m_b\max_{j \in J_{\varepsilon,\kappa}} \nu_j}{n\kappa\min_{i,j}K_{ij}} + (m-m_b)\varepsilon^2 \Big)
\end{equation*}
and 
\begin{equation*}
\norm{\nu^{\text{sc}}}_1 \leq \kappa^{-1} \norm{\nu_{J_{\varepsilon,\kappa}}}_1 + (m-m_b)\Big(\frac{n_b\kappa \max_{i\in I_{\varepsilon,\kappa}}\mu_i}{m\min_{i,j}K_{ij}} + (n-n_b)\varepsilon^2\Big).
\end{equation*}
\end{lemma}
The following Proposition gives also an upper bound of the marginal errors.
\begin{proposition}
\label{proposition_error_in_marginals}
One has 
\begin{align*} %{\norm{{\mu}^{\text{sc}}}_1} 
{\norm{{\mu} -{\mu}^{\text{sc}}}^2_1}&\leq 7n_b(\kappa-\log(\kappa)-1)\max_{i} \mu_i + 7 (n-n_b)\bigg(\frac{m_b\max_{j}\nu_j}{n\kappa\min_{i,j} K_{ij}} + (m- m_b) \varepsilon^2 - \min_{i}\mu_i\\
&+ \max_{i} \mu_i\log\Big(\frac{\kappa(n-n_b+ 1)\max_{i} \mu_i}{m_b\min_{i,j}K_{ij}\min_{j \in J_{\varepsilon,\kappa}}\nu_j} + \frac{\kappa^2\max_{i} \mu_i}{mm_b\varepsilon^2(\min_{i,j}K_{ij})^2\min_{j \in J_{\varepsilon,\kappa}}\nu_j}\Big)% \bigg\}^{1/2}
\end{align*}
and 
\begin{align*} %{\norm{{\nu}^{\text{sc}}}_1}
{\norm{{\nu} -{\nu}^{\text{sc}}}^2_1} &\leq 7m_b(\kappa-\log(\kappa)-1)\max_{j} \nu_j + 7(m-m_b)\bigg(\frac{n_b\kappa\max_{i}\mu_i}{n\min_{i,j} K_{ij}} + (n- n_b) \varepsilon^2 - \min_{j}\nu_j\\
&+ \max_{j} \nu_j\log\Big(\frac{\kappa(m-m_b+ 1)\max_{j} \nu_j}{n_b\min_{i,j}K_{ij}\min_{i \in I_{\varepsilon,\kappa}}\mu_i} + \frac{\kappa^2\max_{j} \nu_j}{nn_b\varepsilon^2(\min_{i,j}K_{ij})^2\min_{i \in I_{\varepsilon,\kappa}}\mu_i}\Big).% \bigg\}^{1/2}
\end{align*}
